Factor analysis

### 2024-11-15 1_get_data.qmd
paper

```{r}
d_sub <- d_trial |>
  group_by(dataset_name, subject_id, administration_id, age) |>
  summarise(n_trials = sum(!is.na(long_window_accuracy)), 
            n_trials_rt = sum(!is.na(rt)),
            rt_var = sd(rt, na.rm=TRUE), 
            log_rt_var = sd(log_rt, na.rm=TRUE), 
            short_window_acc_var = sd(short_window_accuracy, na.rm=TRUE),
            long_window_acc_var = sd(long_window_accuracy, na.rm=TRUE),
            across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
```

```{r}
ggplot(d_sub, aes(x = n_trials, fill = n_trials < 4)) + 
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name, scales = "free_y")
```
```{r}
ggplot(d_sub, aes(x = n_trials_rt, fill = n_trials_rt < 2)) + 
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name, scales = "free_y")
```

@@

```{r}
d_sub <- d_sub |>
  mutate(across(contains("log_rt"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
         across(contains("window"), ~ifelse(n_trials < ACC_MIN_TRIAL_CUTOFF, NA, .x)))

@@

This is very slightly broken due to some repeated fields. We use `values_fn = mean` to suppress them.

@@

d_cdi <- cdi_data |>
  select(administration_id, native_language, measure, language, 
         CDI_percent) |> # rawscore is form-dependent
  filter(language == "English (American)", native_language == "eng") |>
  pivot_wider(names_from = "measure", values_from = "CDI_percent", 
              values_fill = NA, values_fn = mean)
```

```{r}
d_sub <- d_sub |>
  left_join(d_cdi) |>
  mutate(log_age = log(age)) |>
  ungroup()
```

@@

```{r}
print(paste(length(unique(d_sub$subject_id)), "children,",
            sum(!is.na(d_sub$prod)), "with CDI data"))
print(paste(length(unique(d_sub$administration_id)), "administrations"))
print(paste(length(unique(d_sub$dataset_name)), "datasets"))
```

>>>>

```{r}
d_sub_nocdi <- d_trial |>
  group_by(dataset_name, subject_id, administration_id, age) |>
  summarise(n_trials = sum(!is.na(long_window_accuracy)), 
            n_trials_rt = sum(!is.na(rt)),
            rt_var = sd(rt, na.rm=TRUE), 
            log_rt_var = sd(log_rt, na.rm=TRUE), 
            short_window_acc_var = sd(short_window_accuracy, na.rm=TRUE),
            long_window_acc_var = sd(long_window_accuracy, na.rm=TRUE),
            across(c(contains("_window"),"rt","log_rt"), ~mean(.x, na.rm=TRUE)))
```

```{r}
ggplot(d_sub_nocdi, aes(x = n_trials, fill = n_trials < 4)) + 
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name, scales = "free_y")
```
```{r}
ggplot(d_sub_nocdi, aes(x = n_trials_rt, fill = n_trials_rt < 2)) + 
  geom_histogram(binwidth = 1) + 
  facet_wrap(~dataset_name, scales = "free_y")
```

@@

```{r}
d_sub_nocdi_exclusions <- d_sub_nocdi |>
  mutate(across(c("rt", "rt_var"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
         across(contains("log_rt"), ~ifelse(n_trials_rt < RT_MIN_TRIAL_CUTOFF, NA, .x)),
         across(contains("window"), ~ifelse(n_trials < ACC_MIN_TRIAL_CUTOFF, NA, .x)))

@@

This join is very tricky and is done based on some careful work in `peekbank-method` to get the English CDIs and fuzzy join them.

@@

d_sub <- d_sub_nocdi_exclusions |>
  left_join(select(cdi_data, -age, -dataset_name), 
            by = c("subject_id","administration_id")) |>
  mutate(log_age = log(age)) |>
  ungroup()
```

@@

```{r}
print(paste(length(unique(d_sub$subject_id)), "children"))
print(paste(length(unique(d_sub$administration_id)), "administrations"))
print(paste(sum(!is.na(d_sub$prod)| !is.na(d_sub$comp)), "administrations have CDI data"))
print(paste(length(unique(d_sub$dataset_name)), "datasets"))
```

### 2024-11-15 4_factor_analysis.qmd
paper

```{r}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                       ~ scale(.x)[,1])) 
```

@@

```{r}
fa.parallel(select(d_sub_mat, -dataset_name), fa = "fa", 
            use = "pairwise.complete.obs")
```

@@

```{r}
fa(select(d_sub_mat, -dataset_name), nfactor = 2,
            use = "pairwise", rotate = "varimax")
```

@@

# Three-factor mode

@@

@@

We can't really handle the missing data here sadly, because there's too much missing prod and comp data.

```{r}
fa_model_reduced <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var"

fit_reduced <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
                   missing='fiml')
summary(fit_reduced, fit.measures=TRUE, standardize=TRUE)
```

@@

```{r}
fit_reduced_grouped <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
                   missing='fiml', 
                   group = "dataset_name")

summary(fit_reduced_grouped, fit.measures=TRUE, standardize=TRUE)
```

We can look at invariance.

```{r}
fit_reduced_grouped_loadings <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
                   missing='fiml', 
                   group = "dataset_name", 
                   group.equal = "loadings")

@@

lavTestLRT(fit_reduced_grouped, fit_reduced_grouped_loadings)
```

@@

## Fernald and Marchman only

```{r}
fm_data <- d_sub_mat_s |>
  filter(dataset_name %in% c("adams_marchman_2018","fmw_2013"))

fa_model <-  "F1  =~ rt + rt_var + long_window_accuracy + long_window_acc_var + prod + comp"

fm_fit <- cfa(fa_model, fm_data, std.lv=TRUE, missing='fiml')

summary(fm_fit, fit.measures=TRUE, standardize=TRUE)
```
```{r}
fa2_model <-  "accuracy =~ long_window_accuracy + long_window_acc_var + prod + comp
               speed =~ rt + rt_var"

fm_fit2 <- cfa(fa2_model, fm_data, std.lv=TRUE, missing='fiml')

summary(fm_fit2, fit.measures=TRUE, standardize=TRUE)
```

>>>>

```{r}
d_sub_mat <- d_sub |>
  ungroup() |>
  select(dataset_name, rt, rt_var, long_window_accuracy, long_window_acc_var, prod, comp, age)

d_sub_mat_s <- d_sub_mat |>
  ungroup() |>
  mutate(across(all_of(c("rt", "rt_var", "long_window_accuracy", 
                         "long_window_acc_var", "prod", "comp")), 
                       ~ age_scale(.x, age))) 
```

@@

```{r}
fa.parallel(select(d_sub_mat, -dataset_name, -age), fa = "fa", 
            use = "pairwise.complete.obs")
```

@@

```{r}
fa(select(d_sub_mat, -dataset_name), nfactor = 2,
            use = "pairwise", rotate = "varimax")
```
Here's what this looks like not forced to be orthogonal.

```{r}
fa(select(d_sub_mat, -dataset_name, -age), nfactor = 2,
            use = "pairwise", rotate = "oblimin")

@@

# Three-factor model

@@

## Age regression

fa3_age_model <- "
# measurement model
vocab =~ prod + comp
accuracy =~ long_window_accuracy + long_window_acc_var
speed =~ rt + rt_var

# regressions
vocab ~ age
accuracy ~ age
speed ~ age
"

fit3_age <- cfa(fa3_age_model, d_sub_mat_s, std.lv=TRUE, missing='fiml')

summary(fit3_age, fit.measures=TRUE, standardize=TRUE)
layout_age = matrix(nrow=5, ncol = 6,
                data = c(NA,NA,"age",NA, NA, NA,
                         NA, NA, NA, NA, NA, NA, 
                         "speed",NA,"accuracy",NA,"vocab",NA,
                         NA, NA, NA, NA, NA, NA, 
                         "rt_var","rt", "long_window_accuracy",
                         "long_window_acc_var", "prod","comp"), 
                byrow = TRUE)

graph_sem(model = fit3_age, text_size = 2, layout = t(layout_age))
```

@@

These mdels generally do not work sadly.

@@

```{r}
fa3_model_l2 <- "
level: 1
  vocab =~ prod + comp
  accuracy =~ long_window_accuracy + long_window_acc_var
  speed =~ rt + rt_var

level: 2
  vocab =~ prod + comp
  accuracy =~ long_window_accuracy + long_window_acc_var
  speed =~ rt + rt_var"

# optimization from https://lavaan.ugent.be/tutorial/multilevel.html
fit3_grouped <- cfa(fa3_model_l2, d_sub_mat_s, std.lv=TRUE, missing='fiml', 
                    cluster = "dataset_name", 
                    optim.method = "em", 
                    em.iter.max = 20000,
                    em.fx.tol = 1e-08, em.dx.tol = 1e-04)

summary(fit3_grouped, fit.measures=TRUE, standardize=TRUE)
```

We can look at invariance. Not implemented due to convergence issues.

```{r}
# fit_reduced_grouped_loadings <- cfa(fa_model_reduced, d_sub_mat_s, std.lv=TRUE, 
#                    missing='fiml', 
#                    group = "dataset_name", 
#                    group.equal = "loadings")

@@

# lavTestLRT(fit_reduced_grouped, fit_reduced_grouped_loadings)
```

@@
